---
title: "The ridge function, matrices, and prediction"
author: "Terry Therneau"
date: Dec 2020
ouptput: pdf_document
---

# The ridge function

The ridge() function was included in the survival package as a test of the
code for penalized models.  As the author of the package, I never actually
use the function myself, and as a consequence it has never received any
polish.  But it does work.  This neglect is simply due to the types of data
that I analyse.

This vignette was created to address a common issue with using the function,
one that arises with some frequency on internet message boards.
Hopefully this a note will provide context and a refeference for a solution.
I will create a dummy data set with 20 random 0/1 variables as a running
example.

```{r, rdata}
set.seed(1954)  # force reproducability
library(survival)

n <- nrow(lung)
snp <- matrix(rbinom(20*n, 1, p=.1), nrow=n)
snpdata <- cbind(lung, data.frame(snp))
dim(snpdata)
```

## The issue.
The ridge funciton makes the most sense when there are a large number of
variables, for instance if one had 100 SNPs. 
Say that you write the call in the obvious way:

```{r, pass1}
cfit1 <- coxph(Surv(time, status) ~ age + sex + ridge(X1, X2, X3, X4, X5, X6,
                     X7, X8, X9, X10, X11, X12, X13, X14, X15, X16, X17, X18, 
		     X19, X20, theta=.1), data=snpdata)
```

The problem with this, of course, is that typing this statement is cumbersome
for 20 SNPs, impossible if there were 500.
One can use `.' in R formulas to stand for ``all the other variables'', but 
this does not work inside a function.
This leads to the first work-around, which is to create the formula
programatically.

```{r, pass2}
xlist <-  paste0("X", 1:20)
myform <- paste("Surv(time, status) ~ age + sex + ridge(",
                paste(xlist, collapse= ", "), ", theta=.1)")
cfit2 <- coxph(formula(myform), data=snpdata)

all.equal(cfit1$loglik, cfit2$loglik)
```

This approach works well up to a point and then fails, once the model statement
grows too long for the internal buffer of the R parser.
The solution is to call the ridge function with a single matrix argument.

```{r, pass3}
cfit3 <- coxph(Surv(time, status) ~ age + sex + ridge(snp, theta=.1),
      	 data= snpdata)
```

This fit has completely ignored the variable X1, X2, ..., X50 found in the
data frame. 
If the SNP data had come to us as a data frame, then we would
create a temporary  matrix
using the as.matrix command, applied to the appropriate columns of said data.

The problem with this approach comes if we want to do prediction using a new
set of data.

```{r, pass4}
newsnp <- matrix(rbinom(20*4, 1, p=.12), nrow=4)
prdata <- data.frame(age= c(50, 65, 48, 70), sex= c(1, 1, 2,1),
                       newsnp)
predict(cfit3, newdata=prdata)
```

We get a failure message that ``variable lengths differ''.  The reason is that
R fitting functions look for their variables first in the data= argument, 
and then
look in the primary working data for any that were not found.  The call to
cfit3 has 3 variables of age, sex, and snp; while the new data frame prdata
has variables of age, sex, X1, X2, ..., X20.  The predict function pulls 4 rows
from prdata (for age and sex), and 50 rows from the global variable snp.
This clearly does not work.

What if we put those variables in the the main data, instead of a dataframe?

```{r, pass5}
prdata <- data.frame(age= c(50, 65, 48, 70), sex= c(1, 1, 2,1),
                       newsnp)
snpsave <- snp
snp <- newsnp
age <- prdata$age
sex <- prdata$sex

new <- predict(cfit3)
length(new)

snp <- snpsave  # restore the status quo
rm(age, sex)
```

This time we have been tripped up by the predict.coxph function.  If there is
no newdata argument, the function assumes the user is asking for predictions
using the original data, and that is what is produced.  There is also the
need to restore data sets that we overwrote.

The most direct way around this is to do the prediction "by hand".  Since
this data set has no factor variables, splines, or other terms that lead
to special coding, it is fairly easy to do the computation.

```{r, pass6}
prmat <- cbind(age= c(50, 65, 48, 70), sex=c(1,1,2,1), newsnp)
drop(prmat %*% coef(cfit3))
#alternate
drop(prmat %*% coef(cfit3)) - sum(coef(cfit3)* cfit3$mean)
```
The predict.coxph function, by default, gives predictions for centered
covariates.  This has its roots in the internals of the coxph code, where
centering is used to avoid overflow of the exp function.
This is entirely optional when doing the prediction ourselves, unless one
wants to match predict.coxph.  (If I could go back in time, centering the 
predictions would be undone; it's effect has been mostly to cause confusion.
Que sera sera.)

Suppose that one still wanted to use the standard predict function, for instance
if the model did include a factor variable, or simply to fit in to the usual
R pattern?
What is needed is a way to store a matrix directly into a data frame.  This
is done regularly by the model.frame function; the code below causes
data.frame to use that behaviour for our matrices of SNP values.

```{r, pass7}
# new data type
ridgemat <- function(x) {
    class(x) <- c("ridgemat", class(x))
    x
}
as.data.frame.ridgemat <- function(x, ...) 
    as.data.frame.model.matrix(as.matrix(x), ...)

snpdata2 <- cbind(lung, snp= ridgemat(snp))
names(snpdata2)

cfit4 <- coxph(Surv(time, status) ~ age + sex + ridge(snp, theta=.1),
      	     data= snpdata2)
     

prdata <- data.frame(age= c(50, 65, 48, 70), sex= c(1, 1, 2,1),
                       snp= ridgemat(newsnp))
predict(cfit4, newdata=prdata)
```

Success!
The downside to this is that the modified dataframe object is known to work
with modeling functions, but is not guarrateed to be compatable with standard
manipulations for data frames, such as merge, subset, etc., or tidyverse
concepts such as a tibble.


